---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# NSTempRFA

<!-- badges: start -->
<!-- Example badges you can add after uploading to GitHub: [![R-CMD-check](https://github.com/yourusername/NSTempRFA/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/yourusername/NSTempRFA/actions) [![CRAN status](https://www.r-pkg.org/badges/version/NSTempRFA)](https://CRAN.R-project.org/package=NSTempRFA) [![License: GPL-3](https://img.shields.io/badge/license-GPL--3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0) --> 
[![R-CMD-check](https://github.com/gabrielblain/NSTempRFA/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/gabrielblain/NSTempRFA/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/gabrielblain/NSTempRFA/graph/badge.svg)](https://app.codecov.io/gh/gabrielblain/NSTempRFA)
<!-- badges: end -->

A package designed to apply the RFA technique to air temperature data under climate change conditions

## Installation

You can install the development version of NSTempRFA from GitHub with:

``` r
# install.packages("devtools")
devtools::install_github("gabrielblain/NSTempRFA")
```
## Theoretical Background

The Regional Frequency Analysis (RFA; Dalrymple, 1960) technique has been widely used to improve the probabilistic assessment of extreme rainfall and flood events worldwide. The fundamental concept behind RFA is to evaluate whether a group of time series from distinct sites can be considered “acceptably homogeneous” (Hosking and Wallis, 1997). If this condition is met, data from all sites can be pooled together, effectively trading space for time, and thus extending the effective data length (Smith et al. 2015).

In its original version, the RFA is valid only when the variable of interest assumes strictly positive values. However, Martins et al. (2022) verified that this technique can can be effectively applied to extreme maximum (Tmax) and minimum (Tmin) air temperature series, which may assume both positive and negative values, by adopting the so-called additive approach.

This additional approach is consistent with the physical characteristics of temperature data. Since the fundamental temperature scale is the Kelvin scale, where 0 K equals -273.15°C (or -459.67°F), both Tmax and Tmin values fall within a relatively narrow range compared to their distance from absolute zero. As a result, the variability in Tmax and Tmin distributions can be treated as largely independent of their sample mean (Martins et al. 2022). Moreover, when expressed in Celsius or Fahrenheit, Tmin values may be negative, further justifying the use of the additive approach.

Although the study by Martins et al. (2022) showed promising results, it did not address the influence of long-term warming trends, which have been documented in nearly all regions of the world (IPCC, 2021). This calls for an extension of the RFA framework to incorporate time-dependent, nonstationary statistical properties. 

This package proposes a new methodology for applying the RFA to extreme Tmax and Tmin data under nonstationary climate conditions.

## Basic Instructions

### **Load the library for examples below**

```{r}
library(NSTempRFA)
```

### Function `Add_Discord()`

### Description
This function calculates the Hosking and Wallis’ discordance measure (discord) using the additive approach proposed in Martins et al. (2022) doi:10.1590/1678-4499.20220061.

### Usage
Add_Discord(dataset)

### Arguments
**dataset:** A numeric matrix with extreme air temperature data from multiple sites. The
first column must contain the years, and the remaining columns contain temperature
data from each site.

### Details
The discordance measures identify sites that are potentially discordant within a regional frequency analysis framework for extreme air temperature series.

### Value
A data.frame with seven columns:\
- Local Site identifier.\
- SampleSize Number of observations for each site.\
- l_1 L-moment ratio (mean).\
- l_2 L-moment ratio (L-scale).\
- t_3 L-moment ratio (L-skewness).\
- t_4 L-moment ratio (L-kurtosis).\
- t_5 L-moment ratio (higher-order L-moment).\
- discord Original discordance statistic indicating potential outlier status.

### Example

```{r example 1}
dataset <- TmaxCPC_SP
Add_Discord(dataset)
```

### Function `Dataset_add()`

### Description

Subtract the sample mean of each column from their corresponding data.

### Usage
dataset_add(dataset)

### Arguments
**dataset:** A numeric matrix with extreme air temperature data from multiple sites. The first column must contain the years, and the remaining columns contain temperature
data from each site.

### Value
A list object with the following elements:\
- Subtracted_data: A matrix of temperature values subtracted by their sample mean.\
- reg_mean: Sample mean for each site.

### Example

```{r example 2}
dataset <- TmaxCPC_SP
Dataset_add(dataset)
```

### Function `Add_Heterogenety()`

### Description

Calculates Hosking and Wallis' heterogeneity measure
using the additive approach proposed in MARTINS et al. (2022).

### Usage
Add_Heterogenety(dataset.add, rho, Ns)

##Arguments
**dataset.add**: A matrix of temperature values subtracted by their sample mean, as
calculated by the `Add_Discord()` function.\
**rho**: A single numeric value (constant) describing the average correlation among the
sites. It must be larger than -1.0 and lower than 1.0.\
**Ns**: Number of simulated groups of series. Default is 100, but at least 500 is recommended.

### Example
```{r example 3}
rho <- 0.51
Ns <- 500
add.data <- Dataset_add(TmaxCPC_SP)
Add_Heterogenety(dataset.add=add.data$add_data,rho = rho,Ns = Ns)
```

### Function `Best_model()`

### Description
Calculates the time-varying parameters of the best fitting GEV model for each site.
See **Methodological Details** for more information on the candidade
GEV-models considered in this package.

###Usage
Best_model(add_data)

### Arguments
**add_data**: A matrix of temperature values subtracted by their sample mean. May be generated by the `dataset_add()` function

### Value
A list object with the following elements:\
- best: The best model among the six candidates.
- atsite.models: The time-varying parameters of the best model.

### Example
```{r example 4}
add.data <- Dataset_add(TmaxCPC_SP)
Best_model(add.data=add.data$add_data)
```

### Function `Reg_par()`

### Description
Calculates the regional parameters for the regional time-varying GEV distribution.

### Usage
Reg_par(best_model)

### Arguments
**best_model**: A 7-column data.frame as that generated by Best_model() function.\
- 1st column is the mu0 parameters\
- 2nd is the mu1 parameters\
- 3rd is the mu2 parameters\
- 4th is the sigma0 parameters\
- 5th is the sigma1 parameters\
- 6th is the shape parameters\
- 7th is the sample sizes.

### Value
A data.frame with the regional parameters of the regional time-varying distribution.

### Example
```{r example 5}
add.data <- Dataset_add(TmaxCPC_SP)
best.parms <- Best_model(add.data=add.data$add_data)
Reg_par(best_model=best.parms$atsite.models)
```

### Function `Reg_parCI()`

### Description
Calculates the 95% confidence intervals of time-varying parameters of the regional GEV distribution. It follows the method proposed by Burn (2003) and O’Brien and Burn (2014). The spatial dependence between sites is preserved.

### Usage
Reg_parCI(add_data, reg_par, max_time, n.boots)

### Arguments
**add_data**: A matrix of temperature values subtracted by their sample mean. May be generated by the `dataset_add()` function.\
**reg_par**: A 6-column and 1-row data.frame as that generated by reg_par(), that is:
- 1st column is the mu0 parameter\
- 2nd is the mu1 parameter\
- 3rd is the mu2 parameter\
- 4th is the sigma0 parameter\
- 5th is the sigma1 parameter\
- 6th is the shape parameter\
**max_time**: A single number describing the number of the year that the time-varying parameters should be calculated. For example, if the users need to calculated the
parameters for the first year max_time is set to 1 and the 30th year max_time is
set to 30.\
**n.boots**: A single number describing the number of copies of the original dataset. Whenever possible n.boots should be set to 999 (its default value), as suggested by
Burn (2003) and O’Brien and Burn (2014).

### Value
A matrix containing the 95% confidence intervals (lower and upper bounds) of the time-varying parameter estimates. 


### Example
```{r example 6}
add.data <- Dataset_add(TmaxCPC_SP)
best.parms <- Best_model(add.data=add.data$add_data)
regional.parms <- Reg_par(best_model=best.parms$atsite.models)
Reg_parCI(add_data=add.data$add_data,
model=2,
reg_par=regional.parms,
n.boots=100)
```

### Function `Site_parCI`

### Description
Calculates the 95% confidence intervals of time-varying parameters of a at-site GEV distribution. It follows the method proposed by Burn (2003).

### Usage
Site_parCI(data_site, site_par, max_time, n.boots)

### Arguments
**data_site**: A vector or single column matrix of temperature values subtracted by its sample mean\
**site_par**: A 6-column and 1-row data.frame. It may be obtained from best.parms()\
• 1st column is the mu0 parameter\
• 2nd is the mu1 parameter\
• 3rd is the mu2 parameter\
• 4th is the sigma0 parameter\
• 5th is the sigma1 parameter\
• 6th is the shape parameter\
**max_time**: A single number describing the number of the year that the time-varying parameters
should be calculated. For example, if the users need to calculated the
parameters for the first year max_time is set to 1 and the 30th year max_time is
set to 30.\
**n.boots**: A single number describing the number of copies of the original data sample.
Whenever possible n.boots should be set to 999 (its default value), as suggested
by Burn (2003).

### Value
A matrix containing the 95% confidence intervals of the time-varying parameter estimates for a at-site model.

### Example
```{r example 7}
 temperatures <- TmaxCPC_SP$Pixel_1
model <- 2
 site_par <- Fit_model(temperatures, model)
 Site_parCI(atsite_temp=temperatures,
            model=model,
            site_par=site_par[1:5],
            n.boots=100)
```

### Function `Add_RegProb`

### Description
Calculates the cumulative probability of extreme air temperature data using the Additional nonstationary RFA approach based on the GEV distribution.

### Usage
Add_RegProb(quantiles, reg_par, site_mean, max_time)

### Arguments
**quantiles**: A numeric vector with no missing data with extreme air temperatures\
**reg_par**: A 6-column and 1-row data.frame as that generated by reg_par(), that is:
- 1st column is the mu0 parameter\
- 2nd is the mu1 parameter\
- 3rd is the mu2 parameter\
- 4th is the sigma0 parameter\
- 5th is the sigma1 parameter\
- 6th is the shape parameter\
**site_mean**: A single number with site’s average value for the air temperature data\
**max_time**: A single number describing the number of the year that the time-varying parameters
should be calculated. For example, if the users need to calculated the
parameters for the first year max_time is set to 1 and the 30th year max_time is
set to 30.

### Value
Cumulative probabilities of extreme air temperature data calculated from
the Additional nonstationary RFA approach.

### Example
```{r example 8}
 quantiles <- c(37.2, 37.5, 37.8, 37.9, 38.0, 38.1, 38.2, 38.5, 39.0)
 add.data <- Dataset_add(TmaxCPC_SP)
 best.parms <- Best_model(add.data=add.data$add_data)
 regional_pars <- Reg_par(best_model = best.parms$atsite.models)
 Add_RegProb(quantiles=quantiles,
                   regional_pars=regional_pars,
                   site_temp=TmaxCPC_SP$Pixel_1,
                   n.year=34)
```

### Function `Add_RegQuant`

### Description
Calculates quantiles estimates from cumulative probabilities of extreme air temperature data using the Additional nonstationary RFA approach based on the GEV distribution.

### Usage
Add_RegQuant(prob, reg_par, site_mean, max_time)

### Arguments
**prob**: A numeric vector with no missing data with the cumulative probabilities (between 0 and 1).\
**reg_par**: A 6-column and 1-row data.frame as that generated by reg_par(), that is:
- 1st column is the mu0 parameter\
- 2nd is the mu1 parameter\
- 3rd is the mu2 parameter\
- 4th is the sigma0 parameter\
- 5th is the sigma1 parameter\
- 6th is the shape parameter\
**site_mean**: A single number with site’s average value for the air temperature data\
**max_time**: A single number describing the number of the year that the time-varying parameters
should be calculated. For example, if the users need to calculated the
parameters for the first year max_time is set to 1 and the 30th year max_time is set to 30.

### Value
Quantile estimates of extreme air temperature data calculated from
the Additional nonstationary RFA approach.

### Example
```{r example 9}
 prob <- c(0.8, 0.85, 0.90, 0.92, 0.93, 0.94, 0.95, 0.97, 0.99)
 add.data <- Dataset_add(TmaxCPC_SP)
 best.parms <- Best_model(add.data=add.data$add_data)
 regional_pars <- Reg_par(best_model = best.parms$atsite.models)
 Add_RegQuant(
   prob = prob,
   regional_pars = regional_pars,
   site_temp = TmaxCPC_SP$Pixel_1,
   n.year = 34
 )
```

## BugReports:

<https://github.com/gabrielblain/NSTempRFA/issues>

## License:

MIT

## Authors:

Gabriel Constantino Blain, Graciela da Rocha Sobierajski, Leticia Lopes
Martins. Maintainer: Gabriel Constantino Blain,
<gabriel.blain@sp.gov.br>

## Acknowledgments:



## References

Burn, D.H., 2003. The use of resampling for estimating confidence intervals for single site and pooled frequency analysis. Hydrol. Sci. J. 48 (1), 25–38. DOI: 10.1623/hysj.48.1.25.43485.

O’Brien, N.L., Burn,H.D. A nonstationary index-flood technique for estimating extreme quantiles for annual maximum streamflow, Journal of Hydrology, v.519, 2014, DOI: 10.1016/j.jhydrol.2014.09.041.

Dalrymple, T. (1960). Flood frequency analysis. Geological Survive Water Supply Paper, 1543-A, 11-51. DOI: 10.3133/wsp1543A

[IPCC] Intergovernmental Panel on Climate Change (2021). Climate Change 2021: The Physical Science Basis. Contribution of Working Group I to the Sixth Assessment Report of the Intergovernmental Panel on Climate Change. Cambridge University Press. Available at: https://www.ipcc.ch/report/ar6/wg1/ Accessed on: Jan. 1, 2025.

Martins, L. L., Souza, J. C., Sobierajski, G. R. and Blain, G. C. (2022). Is it possible to apply the regional frequency analysis to daily extreme air temperature data? Bragantia, 81, 1-22. DOI: 10.1590/1678-4499.20220061

Smith, A., Sampson, C., & Bates, P. D. (2015). Regional flood frequency analysis at the global scale. Water Resources Research, 51(1), 539–553. DOI: 10.1002/2014WR015814

Hosking JRM, Wallis JR (1997) Regional frequency analysis: an approach based on L-moments. Cambridge University Press. DOI: 10.1017/cbo9780511529443
